---
output:
  md_document
editor_options: 
  markdown: 
    wrap: 72
---

# SEAI 2022 - R - Lab 2

# Intro to R

Vincenzo Nardelli -
[vincnardelli\@gmail.com](mailto:vincnardelli@gmail.com){.email} -
<https://github.com/vincnardelli>

## Lab structure

-   Intro
-   Matrix computation
-   Data manipulation with dplyr
-   Graphs with ggplot2

Let's start from the basics!

```{r}
3 + 5
```

```{r}
12/7
```

```{r}
result <- 3 + 5
result
```

```{r}
print(result)
```

```{r}
result <- result * 3.1415
print(result)
```

```{r}
vector = c(1, 3, 8, 13)
vector
```

Unlike Python, the basic version of R allow operations between scalars
and matrices without loading any external packages.

```{r}
vector[1]
```

The subset is very similar to Python but the index starts from 1 instead
of 0.

Furthermore, in the case of multiple selection, the index starts from 1
and the second value representing the last element is included in the
subset (unlike Python which is NOT included).

```{r}
vector[1:3]
```

```{r}
vector[c(FALSE, TRUE, TRUE, FALSE)]
```

```{r}
vector < 3
```

```{r}
vector[vector < 3]
```

```{r}
1:10
```

```{r}
seq(from=1, to=10, by=1)
```

Inside a vector we can insert just homogeneous class object.

```{r}
class("a")
```

```{r}
class(1)
```

```{r}
c(1, "a")
```

```{r}
class(c(1, "a"))
```

# 1 - Matrix computation

```{r}
a <- matrix(c(10, 8, 5, 12), nrow=2, ncol=2, byrow=TRUE)
a
```

```{r}
dim(a)
```

Subsetting in two dimension

```{r}
a[1, ]
```

```{r}
a[, 1]
```

```{r}
a[1, 1]
```

Operations with matrices

```{r}
a + 2
```

```{r}
a * 2
```

```{r}
t(a)
```

```{r}
b <- matrix(c(5, 3, 15, 6), ncol = 2, byrow = TRUE)
b
```

```{r}
a + b
```

Element-wise multiplication

```{r}
a * b 
```

Matrix multiplication (matricial multiplication)

```{r}
a %*% b
```

```{r}
crossprod(a,b)
```

```{r}
t(a) %*% b
```

Kronecker product

```{r}
a %x% b
```

Determinant

```{r}
det(a)
```

Inverse of a matrix

```{r}
solve(a)
```

# 2 - Data manipulation with dplyr

Let's start loading same data from R.

```{r}
data(mtcars)
mtcars
```

```{r}
class(mtcars)
```

A `data.frame` is an analog of a matrix which can contains a difference
classes of objects for each column. It is the perfect class to store the
data which we commonly use for our analysis.

```{r}
dim(mtcars)
```

```{r}
colnames(mtcars)
```

```{r}
summary(mtcars)
```

Subset a dataframe with base R

```{r}
mtcars[1,]
```

```{r}
mtcars[1:3, ]
```

```{r}
mtcars[, 1]
```

```{r}
mtcars$hp
```

## Data loading

```{r}
path = "https://raw.githubusercontent.com/pandas-dev/pandas/master/doc/data/titanic.csv"
titanic = read.csv(path)
head(titanic)
```

Now we will explore the use of dplyr in subsetting, manipulating and
summarising our data frame. Most of the task that we will cover are
possible also in base R without the use of others packages but dplyr
simplify a lot the work.

## R packages

How to install an R package from CRAN

```{r}
#install.packages("dplyr")
```

Load package in the environment

```{r}
library(dplyr)
```

## Subset variables (columns)

```{r}
select(titanic, PassengerId, Age)
```

## Subset observations (Rows)

```{r}
filter(titanic, Age < 18)
```

## The pipe function

The pipe function %\>% helps to create a series of function to perform
the data wrangling and makes the code more readable. For example if we
want to select some columns and filter some rows we can write...

```{r}
filter(select(titanic, PassengerId, Age), Age <18)
```

As you can see the select function is nested in the filter function to
keep the order we decided. Using pipes the code could be writed as:

```{r}
titanic %>% 
  select(PassengerId, Age) %>% 
  filter(Age < 18)
```

You can think a pipe as the command "then". The previous algorithm can
be interpreted as: - Load data then - Select Id and Age then - Filter
Age less then 18

## Make new variables

```{r}
titanic %>% 
  select(PassengerId, Age) %>% 
  filter(!is.na(Age)) %>% 
  mutate(Age_Bracket = ifelse(Age < 18, 'Minor','Major'))
```

# Arrange data

```{r}
titanic %>% 
  select(PassengerId, Age) %>% 
  filter(!is.na(Age)) %>% 
  mutate(Age_Bracket = ifelse(Age < 18, 'Minor','Major')) %>% 
  arrange(Age)
```

## Group and summarise data

```{r}
titanic %>% 
  select(PassengerId, Survived, Sex) %>% 
  group_by(Sex) %>% 
  summarise(survived_perc = mean(Survived))
```

```{r}
titanic %>% 
  select(PassengerId, Survived, Age) %>% 
  filter(!is.na(Age)) %>% 
  mutate(Age_Bracket = ifelse(Age < 18, 'Minor','Major')) %>% 
  group_by(Age_Bracket) %>% 
  summarise(survived_perc = mean(Survived))
```

```{r}
titanic %>% 
  select(PassengerId, Survived, Age, Sex) %>% 
  filter(!is.na(Age)) %>% 
  mutate(Age_Bracket = ifelse(Age < 18, 'Minor','Major')) %>% 
  group_by(Age_Bracket, Sex) %>% 
  summarise(survived_perc = mean(Survived))
```

```{r}
titanic %>% 
  filter(Sex == 'male') %>%
  group_by(Embarked) %>%
  count()
```

# 2 - Data visualization with ggplot2

```{r}
#install.packages("ggplot2")
library(ggplot2)
```

## Geom bar

```{r}
ggplot(titanic, aes(x=Pclass, fill=Survived)) + 
  geom_bar() +
  labs( y="Number of Passengers", x="Passenger Class", title="Titanic Survival Rate by Passenger Class")
```

```{r}
titanic <- mutate(titanic, Survived = as.factor(Survived))
ggplot(titanic, aes(x=Pclass, fill=Survived)) + 
  geom_bar() +
  labs( y="Number of Passengers", x="Ticket Class", title="Titanic Survival Rate by Passenger Class")
```

```{r}
ggplot(titanic, aes(x=Sex, fill=Survived)) +
  geom_bar() +
  labs( y="Number of Passengers", title="Titanic Survival Rate by Gender by Passenger Class")+ 
  facet_wrap(~Pclass)
```

```{r}
ggplot(titanic, aes(x=Age, fill=Survived))+
  geom_histogram(bins=20) +
  labs(title="Survival Rate by Gender", y="Number of passengers", subtitle = "Distribution by age")
```

```{r}
ggplot(titanic, aes(x=Age, fill=Survived))+
  geom_histogram(bins=20) +
  labs(title="Survival Rate by Gender", y="Number of passengers", subtitle = "Distribution by age, gender and ticket class")+
  facet_grid(Sex~Pclass, scales="free")
```

```{r}
titanic <- mutate(titanic, Pclass = as.factor(Pclass))
 ggplot(data = titanic, aes(x=Pclass, y=Age)) +
  geom_boxplot(alpha=0.7)
```

```{r}
 ggplot(data = titanic, aes(x=Age, y=Fare)) +
  geom_point()
```

```{r}
 ggplot(data = titanic, aes(x=Age, y=Fare, color=Pclass)) +
  geom_point()
```
