---
output:
  md_document
editor_options: 
  markdown: 
    wrap: 72
---

# SEAI 2022 - R - Extra - rgeoda

Vincenzo Nardelli -
[vincnardelli\@gmail.com](mailto:vincnardelli@gmail.com){.email} -
<https://github.com/vincnardelli>

Reference: Xun Li - Geoda Tutorial [Link](https://geodacenter.github.io/rgeoda/articles/rgeoda_tutorial.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`rgeoda` is an R library for spatial data analysis. It is an R wrapper of the libgeoda C++ library, which is built based on the `GeoDa` software. The version used in this tutorial is version 0.0.8.


## 1. Install `rgeoda`

The rgeoda package can be installed using "install.packages()" command:
```
install.packages("rgeoda")
```

, and then can be loaded using the customary "library()" command:

```{r}
library(rgeoda)
```

In addition, the package sf needs to be loaded, since it is a dependency:

```{r}
library(sf)
```

## 2. Load Spatial Data

The rgeoda package for R relies on the sf (simple features) package for basic spatial data
handling functions. In a typical R workflow, one first reads a shape file or other GIS format file with the data using
the sf st_read(file path) command. For example, to load the ESRI Shapefile `Guerry.shp` comes with the package:

```{r}
guerry_path <- system.file("extdata", "Guerry.shp", package = "rgeoda")
guerry <- st_read(guerry_path)
```

Once the spatial object has been created, it can be used to compute a spatial weights matrix
using one of the several weights functions in rgeoda.

## 3. Spatial Weights

Spatial weights are central components in spatial data analysis. The spatial weights represent the possible spatial interactions between observations in space. `rgeoda` provides 6 functions to create 4 different types of spatial weights:

* Contiguity Based Weights: `queen_weights()`, `rook_weights()`
* Distance Based Weights: `distance_weights()`
* K-Nearest Neighbor Weights: `knn_weights()`
* Kernel Weights: `distance_weights()` and `knn_weights()` with kernel parameters

### 3.1 Queen Contiguity Weights

Contiguity means that two spatial units share a common border of non-zero length. Operationally, we can further distinguish between a rook and a queen criterion of contiguity, in analogy to the moves allowed for the such-named pieces on a chess board. The queen criterion is somewhat more encompassing and defines neighbors as spatial units sharing a common edge or a common vertex.

To create a Queen contiguity weights, one can call the function 
```r
queen_weights(sf_obj, order=1, include_lower_order = False, precision_threshold = 0)
``` 
For example, to create a Queen contiguity weights using the sf object `guerry`:

```{r}
queen_w <- queen_weights(guerry)
summary(queen_w)
```

The function `queen_weights()` returns an instance of 
`Weight` object. One can access the meta data of the spatial
weights by accessing the attributes of `GeoDaWeight` object:

#### Attributes of `Weight` object

```{r}
is_symmetric(queen_w)
```

```{r}
has_isolates(queen_w)
```

```{r}
weights_sparsity(queen_w)
```

To access the details of the weights: e.g. list the neighbors of a specified observation:

```{r}
nbrs <- get_neighbors(queen_w, idx = 1)
cat("\nNeighbors of the 1-st observation are:", nbrs)
```
To compute the spatial lag of a specified observation by passing the values of the selected variable:

```{r}
lag <- spatial_lag(queen_w, guerry['Crm_prs'])
lag
```

### 3.2 Rook Contiguity Weights

The rook criterion defines neighbors by the existence of a common edge between two spatial units.
To create a Rook contiguity weights, one can call function: 
```r
rook_weights(sf_obj, order=1,include_lower_order=False, precision_threshold = 0)
``` 
For example, to create a Rook contiguity weights using the sf object `guerry`:

```{r}
rook_w <- rook_weights(guerry)
summary(rook_w)
```

The weights we created are in memory. To save the weights to a file, one can call the function:
```r
save_weights(gda_w, id_variable, out_path, layer_name = "")
```

The `id_variable` defines the unique value of each observation when saving a weights file

The `layer_name` is the layer name of loaded dataset. For a ESRI shapefile, the layer name is the file name without the suffix (e.g. Guerry). 

For example, using Guerry dataset, the column "CODE_DE" can be used as a key to save a weights file:

```{r}
save_weights(rook_w, guerry['CODE_DE'], out_path = '/Users/xun/Downloads/Guerry_r.gal', 
             layer_name = 'Guerry')
```


### 3.3 Distance Based Weights

The most straightforward spatial weights matrix constructed from a distance measure is obtained when i and j are considered neighbors whenever j falls within a critical distance band from i. In order to start the distance based neighbors, we first need to compute a threshold value. `rgeoda` provides a function `min_distthreshold` to help you find a optimized distance threshold that guarantees that every observation has at least one neighbor:

```r
min_distthreshold(GeoDa gda, bool is_arc = False, is_mile = True)
To create a Distance based weights, one can call the function `distance_weights`:
```

Then, with this distance threshold, we can create a distance-band weights using the function:
```r
distance_weights(geoda_obj, dist_thres, power=1.0,  is_inverse=False, is_arc=False, is_mile=True)
``` 

For example:   

```{r}
dist_thres <- min_distthreshold(guerry)
dist_thres
dist_w <- distance_weights(guerry, dist_thres)
summary(dist_w)
```
### 3.4 K-Nearest Neighbor Weights

A special case of distance based weights is K-Nearest neighbor weights, in which every obersvation will have exactly k neighbors. It can be used to avoid the problem of isolate in distance-band weights when a smaller cut-off distance is used. To create a KNN weights, we can call the function `knn_weights`:

```r
knn_weights(gda, k, power = 1.0,is_inverse = False, is_arc = False, is_mile = True)
```

For example, to create a 6-nearest neighbor weights using Guerry:
```{r}
knn6_w <- knn_weights(guerry, 6)
summary(knn6_w)
```

### 3.5 Kernel Weights

Kernel weights apply kernel function to determine the distance decay in the derived continuous weights kernel. The kernel weights are defined as a function K(z) of the ratio between the distance dij from i to j, and the bandwidth hi, with z=dij/hi. 

The kernel functions include

* triangular
* uniform 
* quadratic
* epanechnikov
* quartic
* gaussian

Two functions are provided in `rgeoda` to create kernel weights.

#### Use `kernel_weights` for Kernel Weights with adaptive bandwidth

To create a kernel weights with fixed bandwith:
```{r}
bandwidth <- min_distthreshold(guerry)
kernel_w <- kernel_weights(guerry, bandwidth, kernel_method = "uniform")
summary(kernel_w)
```

The arguments `is_inverse`, `power`, `is_arc` and `is_mile` are the same with the distance based weights. Additionally, `kernel_weights` has another argument that user can specify:
```
use_kernel_diagonals	
(optional) FALSE (default) or TRUE, apply kernel on the diagonal of weights matrix
```

#### Use `kernel_knn_weights` for Kernel Weights with adaptive bandwidth

To create a kernel weights with adaptive bandwidth or using max Knn distance as bandwidth:
```{r}
adptkernel_w = kernel_knn_weights(guerry, 6, "uniform")

summary(adptkernel_w)
```

This kernel weights function two more arguments that user can specify:
```
adaptive_bandwidth	
(optional) TRUE (default) or FALSE: TRUE use adaptive bandwidth calculated using distance of k-nearest neithbors, FALSE use max distance of all observation to their k-nearest neighbors

use_kernel_diagonals	
(optional) FALSE (default) or TRUE, apply kernel on the diagonal of weights matrix
```

## 4 Local Indicators of Spatial Association–LISA


`rgeoda` provides following methods for local spatial autocorrelation statistics:


* Local Moran: local_moran(), local_moran_eb()
* Local Geary: local_geary(), local_multigeary()
* Local Getis-Ord statistics: local_g() and local_gstar()
* Local Join Count: local_joincount(), local_bijoincount(), local_multijoincount()
* Quantile LISA: local_quantilelisa(), local_multiquantilelisa()
* Local Neighbor Match Test: neighbor_match_test()

For more information about the local spatial autocorrelation statisticis, please read Dr. Luc Anselin’s lab notes: http://geodacenter.github.io/workbook/6a_local_auto/lab6a.html.

### 4.1 Local Moran

The Local Moran statistic is a method to identify local clusters and local spatial outliers. For example, we can call  the function `local_moran()` with the created Queen weights and the data “crm_prp = guerry[‘Crm_prp’]” as input parameters:

```{r}
crm_prp = guerry["Crm_prp"]
lisa <- local_moran(queen_w, crm_prp)
```

The `local_moran()` function will return a `lisa` object, and we can access its values/results of lisa computation using the following functions:

* lisa_clusters(): Get the local cluster indicators returned from LISA computation.
* lisa_colors(): Get the cluster colors of LISA computation.
* lisa_labels(): Get the cluster labels of LISA computation.
* lisa_values(): Get the local spatial autocorrelation values returned from LISA computation.
* lisa_num_nbrs(): Get the number of neighbors of every observations in LISA computation.
* lisa_pvalues(): Get the local pseudo-p values of significance returned from LISA computation.
* lisa_fdr(): Get the False Discovery Rate (FDR) in LISA.
* lisa_bo(): Get the False Discovery Rate (FDR) in LISA.

For example, we can call the function `lisa_values()` to get the values of the local Moran:

```{r}
lms <- lisa_values(gda_lisa = lisa)
lms
```

To get the pseudo-p values of significance of local Moran computation:

```{r}
pvals <- lisa_pvalues(lisa)
pvals
```

To get the cluster indicators of local Moran computation:


```{r}
cats <- lisa_clusters(lisa, cutoff = 0.05)
cats
```

The predefined values of the indicators of LISA cluster are:
```
0 Not significant
1 High-High
2 Low-Low
3 High-Low
4 Low-High
5 Undefined
6 Isolated
```
which can be accessed via the function `lisa_labels()`:
```{r}
lbls <- lisa_labels(lisa)
lbls
```

By default, the `local_moran()` function will run with some default parameters, e.g.:
```
significance_cutoff: 0.05
permutation: 999
permutation_method: 'complete'
cpu_threads: 6
seed (for random number generator): 123456789
```
, which are identical to GeoDa desktop software so to replicate the results in GeoDa software. You can set different values when calling the lisa functions.

For example, re-run the above local Moran example using 9,999 permutations. 
```{r}
lisa <- local_moran(queen_w, crm_prp, permutations = 9999)
```
Then, we can use the same `lisa` object to get the new results after 9,999 permutations:
```{r}
pvals <- lisa_pvalues(lisa)
pvals
```

`rgeoda` uses `GeoDa` C++ code, in which multi-threading is used to accelerate the computation of LISA. We can use the argument `ncpu` to specify how many threads to run the computation:

```{r}
lisa <- local_moran(queen_w, crm_prp, cpu_threads = 4)
```

Get the False Discovery Rate value based on current pseudo-p values:
```{r}
fdr <- lisa_fdr(lisa, 0.05)
fdr
```

Then, one can set the FDR value as the cutoff p-value to filter the cluster results:
```{r}
cat_fdr <- lisa_clusters(lisa, cutoff = fdr)
cat_fdr
```

### 4.2 Local Geary

Local Geary is a type of LISA that focuses on squared differences/dissimilarity. A small value of the local geary statistics suggest positive spatial autocorrelation, whereas large values suggest negative spatial autocorrelation. For more details, please read: http://geodacenter.github.io/workbook/6b_local_adv/lab6b.html#local-geary


For example, we can call the function local_geary() with the created Queen weights and the data “crm_prp” as input parameters:


```{r}
geary_crmprp <- local_geary(queen_w, crm_prp)
```

To get the cluster indicators of the local Geary computation:
```{r}
lisa_clusters(geary_crmprp)
```

To get the pseudo-p values of the local Geary computation:

```{r}
lisa_pvalues(geary_crmprp)
```


## Exploratory Spatial Data Analysis

For exploratory spatial data analysis (ESDA), rgeoa provides some utility functions to allow users to easily work with sf to visualize the results and do exploratory spatial data analysis.

###  Start from `sf` package

The sf package has been popular tool to handle geospatial data. It is a good substitue of sp package which will be deprecated soon.

For example, we can simply call plot() function to render the first 9 chorepleth maps using the frist 9 variables in the dataset:

```{r, class.source='rCode',fig.width = 6, fig.height=6}
plot(guerry)
```

### 6.2 ESDA with rgeoda

Now, with the sf object `guerry`, you can call rgeoda's spatial analysis functions. For example, to examine the local Moran statistics of variable "crm_prs" (Population per Crime against persons):

```{r, class.source='rCode'}
queen_w <- queen_weights(guerry)
lisa <- local_moran(queen_w,  guerry['Crm_prs'])
```

Note: rgeoda uses wkb, which is a binary representation of geometries, to exchange data between sf and libgeoda in memory.

###  Create Local Moran Map

With the LISA results, we can make a local moran cluster map:

```{r, class.source='rCode', fig.width = 6, fig.height=6}
lisa_colors <- lisa_colors(lisa)
lisa_labels <- lisa_labels(lisa)
lisa_clusters <- lisa_clusters(lisa)

plot(st_geometry(guerry), 
     col=sapply(lisa_clusters, function(x){return(lisa_colors[[x+1]])}), 
     border = "#333333", lwd=0.2)
title(main = "Local Moran Map of Crm_prs")
legend('bottomleft', legend = lisa_labels, fill = lisa_colors, border = "#eeeeee")
```

In the above code, we use th values of cluster indicators from `rgeoda`'s `LISA` object are used to make the LISA map. We can save the clusters back to the original `sf` data.frame:

```{r}
guerry['moran_cluster'] <- lisa_clusters

```

Checking the values of the cluster indicators, we will see they are integer numbers 0 (not significant), 1 (high-high cluster), 2 (low-low cluster), 3 (low-high cluster), 4 (high-low cluster), 5 (neighborless/island), 6 (undefined):


```{r}
lisa_clusters

```

To create a significance map that is associated with the local Moran map, we can do the same as making the local moran cluster map using the results from lisa_pvalues():

```{r, fig.width = 6, fig.height=6}
lisa_p <- lisa_pvalues(lisa)
p_labels <- c("Not significant", "p <= 0.05", "p <= 0.01", "p <= 0.001")
p_colors <- c("#eeeeee", "#84f576", "#53c53c", "#348124")
plot(st_geometry(guerry), 
     col=sapply(lisa_p, function(x){
       if (x <= 0.001) return(p_colors[4])
       else if (x <= 0.01) return(p_colors[3])
       else if (x <= 0.05) return (p_colors[2])
       else return(p_colors[1])
       }), 
     border = "#333333", lwd=0.2)
title(main = "Local Moran Map of Crm_prs")
legend('bottomleft', legend = p_labels, fill = p_colors, border = "#eeeeee")
```
